Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-05-30 doc/invt.tex
small changes

\documentclass[reprint, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}

@@

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for adaptive flat-distribution-sampling methods,
  such as the Wang-Landau (WL) algorithm and metadynamics.
  %
  These sampling methods
  construct a bias potential to offset
  the potential of mean force
  by incremental updating loops.
  %
  The updating magnitude should decrease over time
  to reduce the fluctuation of the bias potential,
  and the optimal schedule optimizes
  the rate of convergence.
  %
  We derive the optimal schedule from
  an equation of motion
  of a free particle with a position-dependent mass,
  %
  and show that
  the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  we show that the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is given by a modified inverse-time formula.
  %
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the task is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a collective variable $z$.
%
However, $z$
often represents a slow variable
with a localized or multi-modal distribution, $p^*(z)$.
%
Thus,
to capture the global shape of the free energy surface
in terms of the potential of mean force (PMF), $-\ln p^*(z)$,
an effective strategy is to sample a flat distribution
along $z$ by introducing a bias potential that
cancels the PMF.
%The negative logarithm,
%$-\ln p^*(z)$,
%defines a free energy,
%or the potential of mean force (PMF).
%%
%Normally, the distribution $p^*(z)$ is
%limited in one or a few narrow $z$ regions.
%%
%Therefore, to capture the global shape
%of the PMF,
%one has to run multiple independent simulations
%under different umbrella or bias potentials
%along $z$\cite{mezei1987, berg1992, lee1993},
%and then piece together the entire curve.
%%
%This solution can be laborious % clumsy in practice
%and it often fails to converge quickly
%for a complex and glassy system,
%because the system can be trapped
%in a local free energy minimum.
%%for a long time.
%
%
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
%wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential on the fly.
%
These techniques employ regular updating loops
to discourage future visits to configurations
with the current $z$ value
by elevating the bias potential there.
%
%The main difference
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
However,
since the frequent updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying
equilibrium sampling dynamics\cite{laio2002},
one often has to gradually decrease
the updating magnitude
to reduce the effect.

The schedule of reducing
the updating magnitude,
therefore,
determines the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is given by the product of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
(with ``time'' being
the number of simulation steps).
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
We derive the optimal schedule
from an equation of motion of a free particle,
and show that the initial updating magnitude
should be set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
the inverse-time formula,
with a modified origin of time.
%
However,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the modified inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help practitioners
plan and optimize FDS simulations.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects on the optimal schedule.
%
%This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model; or the temperature index
in a simulated tempering simulation\cite{
marinari1992, lyubartsev1992}.
%
For a continuous quantity, $z$,
which is typical in a molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \Delta z)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

For a large system,
the distribution $p^*_i$ tends to
be localized,
%
and to find out the global shape
of the PMF, $-\ln p^*_i$,
it is often advantageous to carry out
a biased sampling that targets
a wider distribution, $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

To do so, we introduce a bias potential, $V_i$,
which changes the equilibrium distribution to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-V_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon the normalization,
$\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i
  =
  \frac{                p^*_i \, e^{-V_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j} }
  .
  \label{eq:pi_p_phi}
\end{equation}
%
Particularly,
to make $\pi_i$ flat,
the bias potential $V_i$
must coincide with $\ln p^*_i$,
up to an additive constant.
%
In practice, $V_i$ may contain error,
and thus $\pi_i$ need not
coincide with the intended distribution $p_i$.

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
an FDS simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL algorithm and metadynamics.
%
These methods update the bias potential
incrementally in regular intervals
and can easily achieve the target distribution.
%
However,
%they are less well understood theoretically, as
the updating loops break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
Thus, in late stages of a long simulation,
the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
if the bias potential were converged to the exact value,
this adaptive simulation tends to the ideal equilibrium simulation
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

For later convenience,
we introduce a shifted bias potential
%
\begin{equation}
  v_i
  \equiv
  V_i
  +
  \ln \frac { p_i }
            { p^*_i }
  .
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
the updates to $v_i$'s are
the same as the updates to $V_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should ideally be a constant of $i$
such that $\pi_i \approx p_i$.

@@

In the original WL papers\cite{
wang2001, wang2001pre},
the updating magnitude $\alpha$
(or $\ln f$ there)
is controlled stage-wise.
%
In each stage, $\alpha$
is kept as a constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, wang2001pre}.
%
This scheme works well for early stages.
%
However, it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $c/t$ with a $c$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.
%
Here we will show that the ideal $c$
is $1.0$, but the origin of $t$ should be shifted
to accommodate the initial error.

As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
several similar ways of
reducing the updating magnitude
were developed for umbrella sampling and metadynamics,
particularly, the self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.
%
Our aim below is to find the optimal schedule $\alpha(t)$,
for a general updating scheme.

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \kappa(t - t')
  .
$
%
Additionally,
we assume that $\kappa(t)$ is an even function and
\begin{equation}
  \lim_{t \rightarrow \infty} \kappa(t)
  = 0
  .
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to
Dirac's $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }.
\label{eq:alpha_invt1}
\end{equation}

@@

We now fix the endpoint value $q(T)$
and variate the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to variating $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{dt} \int_0^T
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
\\
&= \int_0^T
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by assuming Eq. \eqref{eq:kappa_limit}
and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $t$,
we must have
%
$
\ddot u\bigl( q(t') \bigr) = 0
,
$
or
\begin{equation}
  \dot u\bigl( q(t') \bigr) = c
  ,
\label{eq:dudt_const}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:u_def},
we get
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$
yields Eq. \eqref{eq:alpha_invt1}.

@@

The matrix $\mathbf w$ is not arbitrary.
%
First, to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$,
the $\dot v_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant independent of $i$,
if $h_j(t)$ were the same as $p_j$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that the elements can be negative.

>>>>

\documentclass[preprint, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for the class of adaptive sampling methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These sampling methods
  construct a bias potential to offset
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to help reduce the noise in the bias potential,
  and the optimal schedule maximizes
  the rate of convergence.
  %
  We derive the optimal schedule from
  an equation of motion
  of a free particle with a position-dependent mass,
  %
  and show that
  the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  we show that the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse of the time or number of sampling steps with a proper shift.
  %
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
often the task is to compute
a distribution, $p^*(z)$,
along a collective variable $z$ by sampling via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
or molecular dynamics\cite{frenkel, karplus2002} (MD) simulations.
%
However, $z$
often represents a slow variable
with a localized or multi-modal distribution.
%
Thus,
to capture the global shape of the free energy surface
in terms of the potential of mean force (PMF), $-\ln p^*(z)$,
an effective strategy is to introduce a bias potential that
cancels the PMF, which is equivalent to sampling a flat distribution
along $z$\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%The negative logarithm,
%$-\ln p^*(z)$,
%defines a free energy,
%or the potential of mean force (PMF).
%%
%Normally, the distribution $p^*(z)$ is
%limited in one or a few narrow $z$ regions.
%%
%Therefore, to capture the global shape
%of the PMF,
%one has to run multiple independent simulations
%under different umbrella or bias potentials
%along $z$\cite{mezei1987, berg1992, lee1993},
%and then piece together the entire curve.
%%
%This solution can be laborious % clumsy in practice
%and it often fails to converge quickly
%for a complex and glassy system,
%because the system can be trapped
%in a local free energy minimum.
%%for a long time.
%
%
%
%A better solution is to artificially flatten
%the target distribution along $z$\cite{mezei1987, berg1992, lee1993,
%wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%%
%This alleviates the above problem of local trapping
%by forcing the system to travel more
%deliberately along $z$,
%which often represents
%a direction of slow motion.
%%
%To achieve a flat distribution, however,
%the method must be able to construct a
%bias potential that cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
These techniques employ regular updating of the bias potential
to discourage future visits to the current configuration
by incrementally elevating the bias potential there.
%
%The main difference
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
However,
frequent updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying
sampling\cite{laio2002},
and near convergence
one  has to gradually decrease
the updating magnitude.

The schedule of reducing
the updating magnitude,
therefore,
determines the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is a function of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
or
the number of simulation steps.
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
We derive the optimal schedule
from an equation of motion of a free particle,
and show that the initial updating magnitude
should be set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
an inverse-time formula,
with a modified origin of time.
%
However,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for extensive sampling or long simulations,
and the modified inverse-time schedule presented is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results allow one to better
plan and optimize FDS simulations.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of FDS
and some known aspects of the optimal schedule.
%
%This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time or extensive sampling limit
in Sec. \ref{sec:cmpschemes}.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model or the temperature index
in a simulated tempering simulation\cite{
marinari1992, lyubartsev1992}.
%
For a continuous quantity, $z$,
which is typical in a molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \Delta z)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

For a large system at temperatures of interest,
the distribution $p^*_i$ often tends to
be localized,
%
and to explore the global shape
of the PMF, $-\ln p^*_i$,
it is often advantageous to carry out
biased sampling that targets
a wider distribution, $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

To do so, we introduce a bias potential, $V_i$,
which changes the equilibrium distribution to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-V_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon the normalization,
$\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i
  =
  \frac{                p^*_i \, e^{-V_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j} }
  .
  \label{eq:pi_p_phi}
\end{equation}
%
In particular,
to make $\pi_i$ flat,
the bias potential $V_i$
must coincide with $\ln p^*_i$,
up to an additive constant.
%
In practice, $V_i$ may be non-optimal or contain error,
and thus $\pi_i$ need not
coincide with the intended distribution $p_i$.

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
an FDS simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potentials at bin $i$, $V_i$,
are often improved
with the histogram of the sampling, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL algorithm and metadynamics.
%
These methods incrementally update the bias potential
and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the updates  break the microscopic reversibility
of the underlying sampling mechanism. Further failure to properly reduce the size of the histogram increments would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to 
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
Thus, in late stages of a long simulation,
the updating becomes smaller and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
if the bias potential were converged to the exact value,
this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

For  convenience,
we introduce a shifted bias potential
%
\begin{equation}
  v_i
  \equiv
  V_i
  +
  \ln \frac { p_i }
            { p^*_i }
  .
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
the updates to $v_i$'s are
the same as the updates to $V_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should ideally be a constant of $i$
such that $\pi_i \approx p_i$.

@@

In the original WL papers\cite{
wang2001, wang2001pre},
the updating magnitude $\alpha$
(or $\ln f$ there)
is controlled in stages.
%
At each stage, $\alpha$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, wang2001pre}.
%
While this scheme works well for early stages,
%
it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality;
the optimal formula for $\alpha(t)$
might be $c/t$ with a $c$
different from $1$\cite{
morozov2007, zhou2008, morozov2009}.
%
Here we will show that the ideal $c$
is unity, but the origin of $t$ should be shifted
to accommodate the initial error.

As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
several similar ways of
reducing the updating magnitude
were developed for umbrella sampling and metadynamics,
particularly, the self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.
%
Our aim here is to find the optimal schedule $\alpha(t)$,
for a general updating scheme.

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \kappa(t - t')
  .
$
%
Additionally,
we assume that $\kappa(t)$ is an even function and
\begin{equation}
  \lim_{t \rightarrow \infty} \kappa(t)
  = 0
  .
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to a
Dirac $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }.
\label{eq:alpha_invt1}
\end{equation}

@@

We now fix the endpoint value $q(T)$
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying $u$
with fixed endpoints,
$u\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{dt} \int_0^T
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
\\
&= \int_0^T
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by assuming Eq. \eqref{eq:kappa_limit}
and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $t$,
we must have
%
$
\ddot u\bigl( q(t') \bigr) = 0
,
$
or
\begin{equation}
  \dot u\bigl( q(t') \bigr) = c
  ,
\label{eq:dudt_const}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:u_def},
we get
$$
e^{ q(t') - q(T) }
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$
yields Eq. \eqref{eq:alpha_invt1}.

@@

The matrix $\mathbf w$ is not arbitrary.
%
First, to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$,
 $\dot v_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant independent of $i$,
if $h_j(t)$ were the same as $p_j$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that the elements can be negative.
